---
Peer Assessment 1  
Activity Monitoring Data Analysis
---
#Peer Assessment 1
##Activity Monitoring Data Analysis

The activity monitoring data is loaded as an R dataset. The dataset includes the number of steps, date, and interval for each activity.
```{r opendata}
data <- read.csv("activity.csv")
originaldata <- data
drop <- is.na(data$steps)
data <- data[!drop,]
data$date <- as.Date(data$date)
originaldata$date<-as.Date(originaldata$date)
```

1) Calculate the total number of steps taken per day
```{r stepscalc}
library(dplyr)
data<-group_by(data,date)
numsteps <- summarise(data,totalsteps = sum(steps))
```

2) Create a histogram of the total number of steps per activity.
```{r stephist}
hist(numsteps$totalsteps,main="Total Steps per Activity",xlab ="Number of Steps")

```

3) The mean and median of the total number of steps taken per day are
```{r stepssum}
summarysteps <- summarise(data,meansteps = mean(steps),mediansteps=median(steps))
as.data.frame(summarysteps)
```

The average daily activity pattern of the 5-minute interval and the steps are determined and plotted as a time series.
```{r intervalcalc}
data<-ungroup(data)
data<-group_by(data,interval)
intercalc <- summarise(data,meansteps = mean(steps))
plot(intercalc$interval,intercalc$meansteps,type="l",main="Average Daily Activity Pattern",xlab="5 minute interval",ylab="Average Number of Steps")
```

The maximum average number of steps is found on the 5 minute interval.
```{r maxsteps}
intercalc2 <- arrange(intercalc,desc(meansteps))
intercalc2 <- head(intercalc2,1)
max<-intercalc2$interval
```

The interval with the maximum number of steps is `r max`.

The total number of missing data in the dataset is determined.
```{r navalues}
numna <- sum(drop)
```
The total number of missing data in the dataset is `r numna`.

For days with missing data, the missing data is replaced with the average number of steps at that interval taken over the entire number of days in the sample set.
```{r replace}
newdata <- originaldata
mergedata<- merge(newdata,intercalc,by="interval",all.x=T)
drop <- is.na(mergedata$steps)
mergedata <- mutate(mergedata[drop,],steps=meansteps)
mergedata <- group_by(mergedata,date)
alldata <- summarise(mergedata,totalsteps=sum(steps),mean2=mean(steps),median=median(steps))
```

